Carrier phase and amplitude estimation for phase shift keying using pilots and data 1: \\ Estimators and simulations

### 2012-08-16 paper1.tex
Added paper1 and and paper2 just to see how this would look split into too.  Should be fine.

@@

%opening
\title{Carrier phase and amplitude estimation for phase shift keying using pilots and data}
\author{Robby McKilliam, Andre Pollok, Bill Cowley 
\thanks{
Supported under the Australian Government’s Australian Space Research Program.
Robby McKilliam, Andre Pollok and Bill Cowley are with the Institute for Telecommunications Research, The University of South Australia, SA, 5095.
}}

@@

The existing literature mostly considers what is called \emph{noncoherent detection} where no pilots symbols exist ($P = \emptyset$ where $\emptyset$ is the empty set).  In the nocoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is the `strip' the modulation from the recieved signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F$, mapping $\reals$ to $\reals$, is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not entirely obvious how pilot symbols should be included, so as to make the estimator coherent. Techniques are available but they are somewhat adhoc~\cite{Cowley_ref_sym_carr_1998}.  This problem does not occur with the least square estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Swelden's considered the noncoherent setting, but we show in Section~\ref{sec:least-squar-estim}~that only a minor modification of Mackenthun's algorithm is required to include pilot symbols. Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly \emph{symbol error rate}, or \emph{bit error rate}.  Whilst estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite approach here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of specific interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated recieved symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent reciever operations, such as decoding.  Consequently, our metric of performance will not be symbol or bit error rate, it will be mean square error (MSE) between complex amplitude $a_0$ and its estimator $\hat{a}$, that is, our metric is $\expect\abs{a_0 - \hat{a}}^2$ where $\expect$ denotes the expected value. It will actually be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\expect\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ and $\expect(\rho_0 - \hat{\rho})^2$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\expect\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ rather than $\expect(\theta_0 - \hat{\theta})^2$ is the appropriate measure of error for the phase parameter.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*, \label{eq:SS}
\end{align}
where $*$ denotes the complex congugate.  Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find the least squares estimator of $a_0$ as a function of $\{d_i, i \in D\}$,
\begin{equation}\label{eq:hata}
\hat{a} = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain the $SS$ conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given a candidate value for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are atmost $(M+1)\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

@@

\begin{lemma}\label{lem:gg2cont}
Let $X = Z e^{j\Theta}$ be a circularly symmetric complex random variable with pdf $\tfrac{1}{2\pi}f_Z(z)$.  Let $R > 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + X$ and with pdf $f(r,\phi)$.  Let
\[
g(\phi) = \int_{0}^\infty r f(r, \phi) dr,
\]
and
\[
g_2(\phi) = \int_{0}^\infty r^2 f(r, \phi) dr.
\]
Then $g$ and $g_2$ are continuous functions from $\reals$ to $\reals$.
\end{lemma}
\begin{IEEEproof}
BLERG
\end{IEEEproof}

@@

\section{Proof of almost sure convergence (Theorem~\ref{thm:consistency}) } \label{sec:proof-almost-sure}

Substituting $\{ \hat{d}_i(\theta), i \in D \}$ from~\eqref{eq:hatdfinxtheta} into~\eqref{eq:SSallparams} we obtain $SS$ conditioned on minimisation with respect to the data symbols,
 \begin{align*}
SS(\rho, \theta) &=\sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} \hat{d_i}(\theta) }^2 \\
&= A - \rho Z(\theta) - \rho Z^*(\theta) + L \rho^2,
\end{align*}
where
\[
Z(\theta)  = \sum_{i \in P} y_i e^{-j\theta} p_i^* + \sum_{i \in D} y_i e^{-j\theta} \hat{d}_i^*(\theta),
\]
and $Z^*(\theta)$ is the conjugate of $Z(\theta)$.  Differentiating with respect to $\rho$ and setting the resulting expression to zero gives the least squares estimator of $\rho_0$ as a function of $\theta$, 
\begin{equation}\label{eq:hatrhoZ}
\hat{\rho} = \frac{Z(\theta) + Z^*(\theta)}{2L} = \frac{1}{L}\Re(Z(\theta)),
\end{equation}
where $\Re(\cdot)$ denotes the real part.  Substituting this expression into $SS(\rho, \theta)$ gives $SS$ conditioned upon minimisation with respect to $\rho$ and the data symbols,
\[
SS(\theta) = A - \frac{1}{L}\Re(Z(\theta))^2.
\]

We are interested in analysing $\hat{\theta}$, the minimiser of $SS(\theta)$, or equivalently, the maximiser of $\abs{\Re(Z(\theta))}$.  Recalling the definition of $R_i$ and $\Phi_i$ from~\eqref{eq:RiandPhii},
\begin{align*}
y_i &= a_0 s_i + w_i \\
&= a_0 s_i \left( 1 + \frac{w_i}{a_0 s_i} \right) \\
&= a_0 s_i R_i e^{j \Phi_i} \\
&= \rho_0 R_i e^{j ( \Phi_i + \theta_0 + \angle{s_i}) }.
\end{align*}
%The $\Phi_i$ can be identified as \emph{circular~random~variables}~\cite{Mardia_directional_statistics,Fisher1993,McKilliam_mean_dir_est_sq_arc_length2010}.  
Recalling the definition of $\hat{d}_i(\theta)$ and $\hat{u}_i(\theta)$ from~\eqref{eq:hatdfinxtheta},
\begin{align*}
\hat{u}_i(\theta) &= \round{\angle{y_i} - \theta} \\
&= \round{\theta_0 + \Phi_i + \angle{s_i} - \theta} \\
&\equiv \round{ \fracpart{\theta_0 - \theta}_{\pi} + \Phi_i + \angle{s_i}} \pmod{2\pi} \\
&= \round{ \lambda + \Phi_i + \angle{s_i} },
\end{align*}
where we put $\lambda = \fracpart{\theta_0 - \theta}_\pi$.  Because $\hat{d}_i^*(\theta) = e^{-j\hat{u}_i(\theta)}$, it follows that, when $i \in D$,
\begin{align}
 y_i e^{-j\theta} \hat{d}_i^*(\theta) &= \rho_0 R_i e^{j(\lambda + \Phi_i + \angle{s_i} - \round{\lambda + \Phi_i + \angle{s_i}})} \nonumber \\
&= \rho_0 R_i e^{j(\lambda + \Phi_i - \round{\lambda + \Phi_i})} \nonumber  \\
&= \rho_0 R_i e^{j\sfracpart{\lambda + \Phi_i}} \label{eq:yethetadhat}
\end{align}
since $\round{x + \angle{s_i}} = \round{x} + \angle{s_i}$ as a result of $\angle{s_i}$ being a multiple of $\tfrac{2\pi}{M}$.  Otherwise, when $i \in P$,  
\[
y_i e^{-j\theta} p_i^* = \rho_0 R_i e^{j(\lambda + \Phi_i)}.
\]
Now,
\[
Z(\theta) = \rho_0 \sum_{i \in P} R_i e^{j(\lambda + \Phi_i)} + \rho_0  \sum_{i \in D} R_i e^{j\sfracpart{\lambda + \Phi_i}},
\]
and
\[
\Re(Z(\theta)) = \rho_0 \sum_{i \in P} R_i \cos(\lambda + \Phi_i) + \rho_0 \sum_{i \in D} R_i \cos\sfracpart{\lambda + \Phi_i}.
\] 
Let 
\begin{equation}\label{eq:GLdefn}
G_L(\lambda) = \frac{1}{\rho_0 L} \Re(Z(\theta))
\end{equation}
and put $\hat{\lambda}_L = \langle\theta_0 - \hat{\theta}\rangle_\pi$.  Since $\hat{\theta}$ is the maximiser of $\abs{\Re(Z(\theta))}$ then  $\hat{\lambda}_L$ is the maximiser of $\abs{G_L(\lambda)}$.  We will show that $\hat{\lambda}_L$ converges almost surely to zero as $L \rightarrow \infty$.  The proof of part 1 of the Theorem~\ref{thm:consistency} follows from this.

Recall the functions $G$, $h_1$ and $h_2$ defined in the statement of Theorem~\ref{thm:consistency}.  Observe that
\[
\expect G_L(\lambda) = \frac{\abs{P}}{L} h_1(\lambda) + \frac{\abs{D}}{L} h_2(\lambda)
\]
and since $\frac{\abs{P}}{L} \rightarrow p$ and $\frac{\abs{D}}{L} \rightarrow d$ as $L \rightarrow \infty$, then
\[
\lim_{L \rightarrow\infty} \expect G_L(\lambda) = G(\lambda) = p h_1(\lambda)   +  d h_2(\lambda).
\]
As is customary, let $(\Omega, \mathcal F, \prob)$ be the probability space over which the random variables $\{w_i\}$ are defined.  Let $A$ be the subset of the sample space $\Omega$ upon which $\sabs{G(\hat{\lambda}_L)} \rightarrow \sabs{G(0)}$ as $L\rightarrow\infty$.  Lemma~\ref{lem:convtoexpGlamL} shows that $\prob\{A\} = 1$.  Let $A'$ be the subset of the sample space upon which $\hat{\lambda}_L \rightarrow 0$ as $L\rightarrow \infty$.  Because $\sabs{G(x)}$ is uniquely maximised at $x=0$, it follows that $\sabs{G(\hat{\lambda}_L)} \rightarrow \sabs{G(0)}$ only if $\hat{\lambda}_L \rightarrow 0$ as $L \rightarrow\infty$. So $A \subseteq A'$ and therefore $\prob\{A'\} \geq \prob\{A\} = 1$.  Part 1 of Theorem~\ref{thm:consistency} follows.

It remains to prove part 2 of the theorem regarding the convergence of the amplitude estimator $\hat{\rho}$.  From~\eqref{eq:hatrhoZ},
\begin{equation}\label{eq:rhoGLZ}
\hat{\rho} = \frac{1}{L}\Re(Z(\hat{\theta})) = \rho_0 G_L(\hat{\lambda}_L).
\end{equation}  
Lemma~\ref{lem:GLtoG0} shows that $G_L(\hat{\lambda}_L)$ converges almost surely to $G(0)$ as $L\rightarrow\infty$, and $\hat{\rho}$ consequently converges almost surely to $\rho_0 G(0)$ as required.

\begin{lemma}\label{lem:convtoexpGlamL} 
$\sabs{G(\hat{\lambda}_L)} \rightarrow \sabs{G(0)}$ almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Since $\sabs{G(x)}$ is uniquely maximised at $x=0$,
\[
0 \leq \sabs{G(0)} - \sabs{G(\hat{\lambda}_L)},
\]
and since $\hat{\lambda}_L$ is the maximiser of $\abs{G_L(x)}$,
\[ 
0 \leq \sabs{G_L(\hat{\lambda}_L)} - \sabs{G_L(0)}.
\]
Thus,
\begin{align*}
0 &\leq \sabs{G(0)} - \sabs{G(\hat{\lambda}_L)} \\ 
& \leq \sabs{G(0)} - \sabs{G(\hat{\lambda}_L)} + \sabs{G_L(\hat{\lambda}_L)} - \sabs{G_L(0)}\\
& \leq |G(0) - G_L(0)| + |G_L(\hat{\lambda}_L) - G(\hat{\lambda}_L)| \\
&\leq 2\sup_{\lambda \in [-\pi, \pi)} \sabs{G_L(\lambda) - G(\lambda)},
\end{align*}
and the last line converges amost surely to zero by Lemma~\ref{lem:uniflawGGL}.
\end{IEEEproof}

% \begin{lemma}\label{lem:Zconunif}
% \[
% \prob\left\{\sup_{\lambda \in [-\pi, \pi)}\abs{G_L(\lambda) - G(\lambda)} > \epsilon \right\} = O(e^{-\epsilon^2 L}).
% \]
% \end{lemma}
% \begin{IEEEproof}
% \end{IEEEproof}

\begin{lemma}\label{lem:GLtoG0}
$G_L(\hat{\lambda}_L) \rightarrow G(0)$ almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
By the triangle inequality,
\[
\sabs{G_L(\hat{\lambda}_L) - G(0)} \leq \sabs{G_L(\hat{\lambda}_L) - G(\hat{\lambda}_L)} + \sabs{G(\hat{\lambda}_L) - G(0)}.
\]
Now $\sabs{G_L(\hat{\lambda}_L) - G(\hat{\lambda}_L)} \rightarrow 0$ as $L \rightarrow \infty$ as a result of Lemma~\ref{lem:uniflawGGL}, and $\sabs{G(\hat{\lambda}_L) - G(0)} \rightarrow 0$ almost surely as $L \rightarrow \infty$ because $G$ is continuous and $\hat{\lambda}_L \rightarrow 0$ almost surely as $L \rightarrow \infty$.
\end{IEEEproof}

\begin{lemma}\label{lem:uniflawGGL}
The function $G_L$ converges almost surely and uniformly on $[-\pi, \pi)$ to the function $G$, that is,
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - G(\lambda)} \rightarrow 0
\] 
almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Put $T_L(\lambda) = \expect G_L(\lambda)$ and write
\begin{align*}
&\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - G(\lambda)} \\
&\leq \sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T_L(\lambda) + T_L(\lambda) - G(\lambda)} \\
&\leq \sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T_L(\lambda)} +  \sup_{\lambda \in [-\pi, \pi)}\sabs{T_L(\lambda) - G(\lambda)}.
\end{align*}
Now,
\begin{align*}
T_L(\lambda) - G(\lambda) &= \big(\tfrac{\abs{P}}{L} - p\big)h_1(\lambda) + \big(\tfrac{\abs{D}}{L} - d\big)h_2(\lambda) \\
&= o(1) h_1(\lambda) + o(1) h_2(\lambda)
\end{align*}
where $o(1)$ denotes a number converging to zero as $L \rightarrow \infty$.  Since 
\[
h_1(\lambda) = \expect R_1\cos(\lambda + \Phi_1) \leq \expect R_1,
\]
and 
\[
h_2(\lambda) = \expect R_1\cos\fracpart{\lambda + \Phi_1} \leq \expect R_1
\] 
for all $\lambda \in [-\pi, \pi)$, it follows that 
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{T_L(\lambda) - G(\lambda)} \leq o(1) \expect R_1 \rightarrow 0
\]  
as $L\rightarrow \infty$.  Lemma~\ref{lem:uniflawTGL} shows that 
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T_L(\lambda)} \rightarrow 0
\]
almost surely as $L\rightarrow \infty$.
\end{IEEEproof}

\begin{lemma}\label{lem:uniflawTGL} 
Put $T_L(\lambda) = \expect G_L(\lambda)$, then
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T_L(\lambda)} \rightarrow 0
\] 
almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Put $D_L(\lambda) = G_L(\lambda) - T_L(\lambda)$ and let
\[
\lambda_n = \tfrac{2\pi}{N}(n-1) - \pi, \qquad n = 1, \dots, N
\]
be $N$ points uniformly spaced on the interval $[-\pi, \pi)$.  Let $L_n$ be the interval $[\lambda_n, \lambda_n + \tfrac{2\pi}{N})$ and observe that $L_1, \dots, L_N$ partition $[-\pi, \pi)$.  Now
\begin{align*}
\sup_{\lambda \in [-\pi, \pi)}&\sabs{D_L(\lambda)} \\
&= \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n) + D_L(\lambda_n)} \\
&\leq U_L + V_L,
\end{align*}
where 
\[
U_L = \sup_{n = 1,\dots,N}\sabs{D_L(\lambda_n)},
\]
\[
V_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n)}.
\]
Lemma~\ref{lem:supDLlambdan} shows that for any finite $N$, and any $\epsilon > 0$, 
\[
\prob\left\{ \lim_{L\rightarrow\infty} U_L > \epsilon \right\} = 0,
\]
i.e., $U_L \rightarrow 0$ almost surely as $L\rightarrow\infty$.  Lemma~\ref{lem:supsupDLlambda} shows that for any $\epsilon > 0$,
\[
\prob\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{4\pi }{N}\expect R_i \right\} = 0.
\] 
Choose $N$ large enough that $4\pi\expect R_i < \epsilon N$ and
\begin{align*}
\prob& \left\{ \lim_{L\rightarrow\infty} \sup_{\lambda \in [-\pi, \pi)}\sabs{D_L(\lambda)} > 3\epsilon \right\} \\
&\leq \prob\left\{\lim_{L\rightarrow\infty}(U_L + V_L) > 3\epsilon \right\} \\
&\leq \prob\left\{\lim_{L\rightarrow\infty} (U_L + V_L) > \epsilon + \epsilon + \tfrac{4\pi }{N}\expect R_i \right\} \\
&\leq \prob\left\{\lim_{L\rightarrow\infty} U_L > \epsilon\right\} +  \prob\left\{ \lim_{L\rightarrow\infty} V_L > \epsilon + \tfrac{4\pi }{N}\expect R_i \right\} \\
&= 0.
\end{align*}
Thus, $\sup_{\lambda \in [-\pi, \pi)}\sabs{D_L(\lambda)} \rightarrow 0$ almost surely as $L\rightarrow\infty$.
\end{IEEEproof}

\begin{lemma}\label{lem:supDLlambdan}
If $N$ is finite then $U_L \rightarrow 0$ almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Put
\[
Z_i(\lambda) = \begin{cases}
R_i\cos(\lambda + \Phi_i) , & i \in P \\
R_i\cos\sfracpart{\lambda + \Phi_i} , & i \in D
\end{cases}
\]
so that
\[
D_L(\lambda) = G_L(\lambda) - T_L(\lambda) = \frac{1}{L}\sum_{i\in P \cup D} \big( Z_i(\lambda) - \expect Z_i(\lambda) \big).
\]
Now $Z_1(\lambda_n), \dots,Z_L(\lambda_n)$ are independent with finite variance (because $\expect R_i^2$ is finite), so for each $n =1, \dots, N$,
\[
\abs{D_L(\lambda_n)} = \abs{\frac{1}{L}\sum_{i\in P \cup D} \big( Z_i(\lambda_n) - \expect Z_i(\lambda)\big)} \rightarrow 0
\]
almost surely as $L\rightarrow\infty$ by the Kolmogorov's strong law of large numbers~\cite[Theorem 2.3.10]{SenSinger_large_sample_stats_1993}.  Now, since $N$ is finite,
\[
U_L = \sup_{n=1,\dots,N}\sabs{D_L(\lambda_n)} \leq \sum_{n=1}^N \sabs{D_L(\lambda_n)} \rightarrow 0
\]
almost surely to zero as $L \rightarrow \infty$.
\end{IEEEproof}

\begin{lemma}\label{lem:supsupDLlambda} For any $\epsilon > 0$,
\[
\prob\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{2\pi}{N}\expect R_i \right\} = 0.
\]
\end{lemma}
\begin{IEEEproof}
Observe that
\begin{align*}
&\sabs{D_L(\lambda) - D_L(\lambda_n)} \\
&=  \abs{G_L(\lambda) - T_L(\lambda) - G_L(\lambda_n) + T_L(\lambda_n)} \\
&\leq  \abs{G_L(\lambda) - G_L(\lambda_n)} + \abs{\expect G_L(\lambda) - \expect G_L(\lambda_n)} \\
&\leq \abs{G_L(\lambda) - G_L(\lambda_n) } + \expect \abs{G_L(\lambda) - G_L(\lambda_n)},
\end{align*}
the last line following from Jensens inequality.  Put
\[
C_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n} \abs{G_L(\lambda) - G_L(\lambda_n) },
\]
so that
\[
V_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n)} \leq C_L + \expect C_L,
\]
because 
\[
 \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n} \expect \abs{G_L(\lambda) - G_L(\lambda_n)} \leq \expect C_L
\]
follows by moving expectation outside the supremum.  Lemma~\ref{lem:CL} shows that $\expect C_L \leq \tfrac{2\pi}{N}\expect R_1$ and also that
\[
\prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\]
So,
\begin{align*}
\prob&\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{4\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} (C_L + \expect C_L) > \epsilon + \tfrac{4\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\end{align*}
\end{IEEEproof}

\begin{lemma}\label{lem:CL} The following statements hold:
\begin{enumerate}
\item $\expect C_L \leq \tfrac{2\pi}{N}\expect R_1$,
\item for any $\epsilon > 0$, $\prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0$.
\end{enumerate}
\end{lemma}
\begin{IEEEproof}
If $\lambda \in L_n$, then $\lambda = \lambda_n + \delta$ with $\delta < \tfrac{2\pi}{N}$, and from Lemma~\ref{lem:coslipshitz},
\begin{align*}
&\abs{\cos(\lambda + \Phi_i) - \cos(\lambda_n + \Phi_i)} \leq \tfrac{2\pi}{N}, \;\;\; \text{and} \\
&\abs{\cos\sfracpart{\lambda + \Phi_i} - \cos\sfracpart{\lambda_n + \Phi_i}} \leq \tfrac{2\pi}{N}.
\end{align*}
Because this results does not depend on $n$,
\[
\sup_{n=1,\dots,N}\sup_{\lambda \in L_n} \abs{Z_i(\lambda) - Z_i(\lambda_n)} \leq R_i \frac{2\pi}{N}
\]
for all $i = P \cup D$.  Now,
\begin{align*}
C_L &= \sup_{n=1,\dots,N}\sup_{\lambda \in L_n}\abs{\frac{1}{L}\sum_{i \in P \cup D} Z_i(\lambda) - Z_i(\lambda_n)} \\
&\leq \frac{1}{L}\sum_{i \in P \cup D} \sup_{n=1,\dots,N}\sup_{\lambda \in L_n} \abs{Z_i(\lambda) - Z_i(\lambda_n)} \\
&\leq \frac{2\pi}{N L}\sum_{i \in P \cup D} R_i .
\end{align*}
Thus, 
\[
\expect C_L \leq \expect \frac{2\pi}{N L}\sum_{i \in P \cup D} R_i = \tfrac{2\pi}{N}\expect R_1
\]
and the first statement holds.  Now, 
\[
\frac{2\pi}{N L}\sum_{i \in P \cup D} R_i \rightarrow \frac{2\pi}{N}\expect R_1
\] 
almost surely to $\frac{2\pi}{N}\expect R_1$ as $L \rightarrow\infty$ by the strong law of large numbers, and so, for any $\epsilon > 0$,
\begin{align*}
\prob&\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} \frac{2\pi}{N L}\sum_{i \in P \cup D} R_i > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\end{align*}
\end{IEEEproof}

>>>>

\usepackage{xr}
\externaldocument{paper2}

@@

%opening
\title{Carrier phase and amplitude estimation for phase shift keying using pilots and data 1: \\ Estimators and simulations}
\author{Robby McKilliam, Andre Pollok, Bill Cowley  
\thanks{
Supported under the Australian Government’s Australian Space Research Program.
Robby McKilliam, Andre Pollok and Bill Cowley are with the Institute for Telecommunications Research, The University of South Australia, SA, 5095.
}}

@@

The existing literature mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$ where $\emptyset$ is the empty set).  In the nocoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is the `strip' the modulation from the recieved signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F$, mapping $\reals$ to $\reals$, is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not entirely obvious how pilot symbols should be included, so as to make the estimator coherent. Techniques are available but they are somewhat adhoc~\cite{Cowley_ref_sym_carr_1998}.  This problem does not occur with the least square estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Swelden's considered only the noncoherent setting, but we show in Section~\ref{sec:least-squar-estim}~that only a minor modification of Mackenthun's algorithm is required to include pilot symbols. Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly \emph{symbol error rate}, or \emph{bit error rate}.  Whilst estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite approach here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated recieved symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent reciever operations, such as decoding.  Consequently, our metric of performance will not be symbol or bit error rate, it will be mean square error (MSE) between complex amplitude $a_0$ and its estimator $\hat{a}$, that is, our metric is $\expect\abs{a_0 - \hat{a}}^2$ where $\expect$ denotes the expected value. It will actually be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\expect\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ and $\expect(\rho_0 - \hat{\rho})^2$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\expect\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ rather than $\expect(\theta_0 - \hat{\theta})^2$ is the appropriate measure of error for the phase parameter.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*, \label{eq:SS}
\end{align}
where $*$ denotes the complex congugate.  Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find the least squares estimator of $a_0$ as a function of $\{d_i, i \in D\}$,
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain the $SS$ conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are atmost $(M+1)\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

@@

\begin{lemma}\label{lem:gg2cont}
Let $X = Z e^{j\Theta}$ be a circularly symmetric complex random variable with pdf $\tfrac{1}{2\pi}f_Z(z)$.  Let $R > 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + X$ and with joint pdf $f(r,\phi)$.  Then
\[
g(\phi) = \int_{0}^\infty r f(r, \phi) dr,
\]
and
\[
g_2(\phi) = \int_{0}^\infty r^2 f(r, \phi) dr.
\]
are continuous functions from $\reals$ to $\reals$.
\end{lemma}
\begin{IEEEproof}
BLERG
\end{IEEEproof}

@@
